---
title: Appendix for Malishev, M & Civitello, DJ Linking bioenergetics and parasite transmission models suggests mismatch between snail host density and production of human schistosomes  
#bibliography:/Users/malishev/Documents/Melbourne Uni/Thesis_2016/library.bib
fontsize: 10
geometry: margin=1in
documentclass: article
linkcolor: blue
urlcolor: blue
citecolor: red
output:
  html_document:
    highlight: tango
    code_folding: show
    toc: no
    toc_depth: 4
    number_sections: no
    toc_float: yes
  sansfont: Times New Roman
  word_document:
    highlight: tango
    keep_md: no
    pandoc_args: --smart
    #reference: mystyles.docx
    toc: yes
  pdf_document:
    includes:
      in_header: # add .tex file with header content
    highlight: tango
    template: null
    toc: no
    toc_depth: 4
    number_sections: false
    fig_width: 4
    fig_height: 5
    fig_caption: true
    df_print: tibble 
    # citation_package: biblatex # natbib
    latex_engine: xelatex #pdflatex # lualatex
    keep_tex: true # keep .tex file in dir 
inludes:
  before_body: before_body.tex
subtitle: 
tags:
- nothing
- nothingness
params: 
  dir: "/Users/malishev/Documents/Emory/research/schisto_ibm/DEB_IBM"
  date: !r Sys.Date()
  session: !r sessionInfo()  
  version: !r getRversion()
  email: "matthew.malishev@gmail.com"
  doi: https://github.com/darwinanddavis/MalishevCivitelloSICB
classoption: portrait
# ^['https://github.com/darwinanddavis/UsefulCode'] # footnote
vignette: >
  %\VignetteIndexEntry{Useful R code}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

<script type="text/x-mathjax-config">
  MathJax.Hub.Config({ TeX: { equationNumbers: {autoNumber: "all"} } });
</script>

```{r echo = FALSE}
# library(rmarkdown)
# setwd("")
# f <- list.files()[1]
# render(f, output_format='pdf_document')
```

## Model construction

We built an individual-based disease transmission and dynamic energy budget model for schistosomiasis infection (SIDEB) of a size-structured *Biomphalaria glabrata* host population and its human infectious parasite *Schistosoma mansoni*. The SIDEB model is informed by two sub models: a complete dynamic energy and mass budget (DEB) model for *B. glabrata*, including energy and mass compartments for *S. mansoni* parasite biomass and a disease transmission model.  

\  
\  
   
Table A1. Key dynamic energy budget state variables for individual hosts and within-host parasite biomass (bold). Equations refer to the equation numbers listed under 'Dynamic energy budget for _B. glabrata_'. Dimensions and units: ---, dimensionless; C, carbon; d, day; L, length; mg, milligrams; mm, millimeters.      

| Parameter | Definition | Dimension<br/>(unit) | Equation |     
| :---: | :--- | :---: | :--- |    
| _L_ | structural length | _mm_ | Eq. 2 |        
| _e_ | scaled reserve density | --- | Eq. 3 |    
| _D_ | host development | _mg C_ | Eq. 4 |      
| $R_H$ | energy in reproduction buffer| _mg C_ | Eq. 5 |    
| _$\delta$_ | damage density (infection and starvation costs) | $d^{-1}$ | Eq. 8 |      
| _H_ | starvation-related hazard rate | $d^{-1}$ | Eq. 9 |    
| $L_{G}$ |  shell length | _L (mm)_ | --- |  
| age |  age of egg (hatches at 10 days) | $d$ | --- |    
| **_P_** |  **parasite biomass** | **_mg C_** | **Eq. 6** |      
| **$R_P$** |  **energy in parasite reproduction** | **_mg C_** | **Eq. 7** |

\  
\  

Table A2. Definitions of model state variables for the simulation model environment. Dimensions and units: C, carbon; g, grams; L, liter; N, number; mg, milligrams.      

| Parameter | Definition | Dimension<br/>(unit) |      
| :---: | :--- | :---: |    
| $F_a$ | algae resource density | _mg C_ $L^{-1} d^{-1}$ |    
| $F_d$ | detritus resource density | _mg_ _C_ $L^{-1} d^{-1}$ |      
| $M$ | total number of miracidia | _N_ $L^{-1}$ |  
| $C$ | cercarial density | _N_ $L^{-1}$ |  
| $G$ | total number of host eggs | _N_ $L^{-1}$ |  

\  
\  

### Dynamic energy budget sub-model for *B. glabrata*.      

We outline the energy and mass budget equations for the dynamic energy and mass budget state variables for the intermediate snail host *Biomphalaria glabrata*. All equations are in [Civitello *et al.* (2018)](Civitello, D. J., Fatima, H. , Johnson, L. R., Nisbet, R. M., Rohr, J. R. and Ben‐Ami, F. (2018), Bioenergetic theory predicts infection dynamics of human schistosomes in intermediate host snails across ecological gradients. Ecol Lett, 21: 692-701). Equation number in brackets refers to the equation references in Civitello *et al.* (2018).      
  
\   
\  

Resource, $F$, growth in the environment (Eq. 1)    

\  
\

$$
  \frac 
    {d F}
    {d t}
    = -i_{M}L^{2}f_{H}
$$

\    
\  
   
where $i_{M}$ is the maximum ingestion rate per individual host, $L$ is host length ($cm$), and $f_H$ is the host functional feeding response.    

\    
\  
     
Host body length, $L$ (growth) (Eq. 2)  

\    
\  
     
$$
  \frac 
    {d L}
    {d t}
    = \frac
      {gY_{VE}}
      {3\chi} 
        \cdot \left( \frac 
        {\kappa^{*} a_{M}e - (m_{V} + m_{R}E_{M}\delta)\chi L}
        {e + g}
        \right)
$$

\    
\  
   
where $g$ is the energy investment ratio per unit of biomass $cm^{3}$, $Y_{VE}$ is yield of structure on host energy reserve, $\chi$ is the ratio of structural biomass to physical length (mg C mm$^{-3}$), $\kappa$ is the allocation of mobilized energy reserve $e$ to somatic maintenance $m_V$ and growth, $a_M$ is the maximum assimilation rate of food converted into energy, $m_R$ is maintenance of infected (damaged) biomass, $E_M$ is maximum reserve (per unit of structural biomass $V$), and $\delta$ is scaled damage density.    
  
\      
\    
     
Scaled host energy reserve, $e$ (Eq. 3)  

\    
\  
     
$$
  \frac 
    {d e}
    {d t}
    = (f_{H} - e)
    \frac
      {a_{M}}
      {\chi E_{M}L} - \frac
        {P} 
        {\chi L^{3}} 
          \left(
          \frac 
          {i_{PM}f_{P}}
          {E_{M}}
          \right)
$$

\    
\  
   
where $P$ is the within-host parasite biomass (growth of parasite), $i_{PM}$ is the maximum mass-specific within-host parasite ingestion rate, and $f_P$ is the parasite functional feeding response.  

\    
   
The $\frac{P} {\chi L^{3}} \left(\frac {i_{PM}f_{P}}{E_{M}}\right)$ term equals the depleted energy reserve of infected hosts by parasites.     

\    
\  
     
Host development, $D$ (Eq. 4)  

\    
\  
     
$$
  \frac
  {d D}
  {d t}
  = \begin{cases}
      \text{if} \ D < D_{R}, & (1 - \kappa^{*})C - m_{D}D \\
      \text{if} \ D \geq D_{R}, & \text{min}[\ 0,\ (1 - \kappa^{*})C - m_{D}D\ ] \\
    \end{cases}
$$

\    
\  
   
where $D_R$ is the maturity threshold for host reproduction, $C$ is the rate of mobilized host reserve, and $m_D$ is the rate of host maturity maintenance.  
  
\  
  
The $\text{min}[\ 0,\ (1 - \kappa^{*})C - m_{D}D\ ] \\$ term allows the host to 'regress' back to a previous maturity stage when volume-specific reserve becomes too low under starvation conditions, i.e. the shrinking and regression starvation rule (Kooijman 2010).      

\    
\  
   
Host reproduction, $R_H$ (Eq. 5)    

\    
\  
     
$$
  \frac
  {d R_{H}}
  {d t}
  = \begin{cases}
      \text{if} \ D < D_{R}, & 0 \\
      \text{if} \ D \geq D_{R}, & (1 - \kappa^{*})C - m_{D}D
    \end{cases}
$$

\    
\  
   
Within-host parasite biomass, $P$ (growth) (Eq. 6)  

\    
\  
     
$$
  \frac
  {d P}
  {d t}
  = P(Y_{PE}i_{PM}f_{P} (1-r_{P}) -m_{P})
$$

\    
\  
   
where $Y_{PE}$ is the yield of parasite biomass on host energy reserve, $r_P$ is the allocated parasite biomass assimilated to new parasite offspring (proportional to parasite biomass), and $m_P$ is the mass-specific rate of parasite maturity maintenance.     

\    
\  

\newpage     
Parasite reproduction, $R_P$ (Eq. 7)  

\    
\  
     
$$
  \frac
  {d R_{P}}
  {d t}
  = \gamma_{RP}Y_{PE}i_{PM}f_{P}r_{P}
$$

\    
\  
   
where $\gamma_{RP}$ is the relative yield of parasite offspring biomass from assimilated parasite reserve.  

\    
\  
   
Host damage density, $\delta$ (Eq. 8)  

\    
\  
     
$$
  \frac 
    {d \delta}
    {d t}
    = \frac
      {\Theta} 
      {\chi L^{3}} 
      \cdot \frac
      {dR_{P}}
      {dt}
        + k_{R}(1-e)-k_{R}\delta-
        \frac
          {3\delta}
          {L}
          \cdot \frac 
            {dL}
            {dt}
$$

\    
\  
   
where $\Theta$ is the intensity of parasite-induced damage to the host and $k_R$ is a constant for the rate of damage repair via reserve depletion. The $\frac {3\delta}{L}\cdot \frac {dL}{dt}$ term is host "dilution by growth": as the host adds new biomass, this reduces the damage density to its overall biomass.  

\    
\  
   
Cumulative hazard for host, $H$ (Eq. 9)  

\    
\  
     
$$
  \frac
  {d H}
  {d t}
  = h_{b}+h_{\delta} \cdot \text{max}(0, \ \delta - \delta_{0})
$$

\    
\  
   
where $h_b$ is the background hazard rate, $h_{\delta}$ is the hazard coefficient, and $\delta_{0}$ is the linear function for damage density.    

\    
\  
   
Host survival probability (Eq. 10)  

\    
\  
     
$$
P(\text{Survival})[t] = e^{-H(t)}
$$
  
\newpage  

## Simulation model code for the within-host DEB model, the between-host disease transmission model, and the individual-based simulation model    

\  
\  

All model code is found at [https://github.com/darwinanddavis/SchistoIBM](https://github.com/darwinanddavis/SchistoIBM).  

\  
\  

Required files  

\  
\ 

```{shell, eval=F, echo=T}
"DEB_IBM.R"
"DEB_INF_GUTS_IBM.nlogo"
"FullStarve_shrink_dilute_damage3.Rda"
"IndividualModel_IBM3.c"
"IndividualModel_IBM3.so" # Mac OSX. generated from C
"IndividualModel_IBM3.o" # Mac OSX. generated from C  
"IndividualModel_IBM.dll" # Windows. generated from C  
```

\  
\  

### Individual-based simulation model code.

\  
\  

```{r, eval=F, echo=T}

#################################  Running NetLogo in Mac ##################################
# if using Mac OSX El Capitan+ and not already in JQR, download and open JGR 
mac <- 0

if(mac==1){
  install.packages('JGR',,'http://www.rforge.net/')
  library(JGR)
  JGR::JGR()
}

#############################################################################################
############################# Windows or JGR onwards ########################################
#############################################################################################

####################################  set user inputs ####################################### 
# isolate sensitive data:
# "FullStarve_shrink_dilute_damage3.Rda"

# set user outputs
snab <- 0 # 1 = use remote access, 0 = run model on your comp 
mac <- 1 # mac or windows system? 1 = mac, 0 = windows 
gui <- 0 # display the gui? 1 = yes, 0 = no
pck <- 0 # if not already, install rnetlogo and rjava from source? 1 = yes, 0 = already installed 
save_to_file <- 0 # 1 = save simulation outputs to local dir, 0 = plot in current R session
mcmcplot <- 0 # 1 = save mcmc plots to dir
traceplot <- 0 # 1 = include traceplot in mcmc plots? intensive!!!

# set dir paths (for "/" for both Windows and Mac)
if(snab==1){
  # set dir paths (for "/" for both Windows and Mac)
  wd <- "R:/CivitelloLab/matt/schisto_ibm" # set working directory  
  ver_nl <-"6.0.4"# type in Netlogo version. found in local dir. 
  ver_gcc <-"4.6.3" # NULL # type in gcc version (if known). leave as "NULL" if unknown   
  nl.path <- "C:/Program Files" # set path to Netlogo program location
}else{
  wd <- "/Users/malishev/Documents/Emory/research/schisto_ibm/SchistoIBM" # set working directory  
  ver_nl <-"6.0.4" # type in Netlogo version. found in local dir. 
  ver_gcc <-"4.6.3" # NULL # type in gcc version (if known). leave as "NULL" if unknown   
  nl.path <- "/Users/malishev/Documents/Melbourne Uni/Programs" # set path to Netlogo program location
}

# define starting conditions for simulation model @netlogo
n.ticks <- 120 # set number of days to simulate
day <- 1 # number of days to run simulation  
resources <- "cyclical" # set resources: "cyclical" or "event"
resource_type <- "algae" # set resource type as "algae" or "detritus"

####################################  set model paths #######################################
# load files
deb_samps <- "FullStarve_shrink_dilute_damage3.Rda"
deb_compile <- "IndividualModel_IBM3"

setwd(wd)
nl.model <- list.files(pattern="*.nlogo") ;nl.model # Netlogo model
if(mac==1){
  nl.path <- paste0(nl.path,"/NetLogo ",ver_nl,"/Java/"); cat("Mac path:",nl.path)
}else{
  nl.path <- paste0(nl.path,"/NetLogo ",ver_nl,"/app/"); cat("Windows path:",nl.path)
}
model.path <- paste0(wd,"/"); model.path # set path to Netlogo model   

####################################  load packages #######################################
# if already loaded, uninstall RNetlogo and rJava
if(pck==1){
  p<-c("rJava", "RNetLogo"); remove.packages(p)
  # then install rJava and RNetLogo from source
  if(mac==1){
    install.packages("rJava", repos = "https://cran.r-project.org/", type="source"); library(rJava)
    install.packages("RNetLogo", repos = "https://cran.r-project.org/", type="source"); library(RNetLogo)
  }
}

# check pck versions
installed.packages()["RNetLogo","Version"] 
installed.packages()["rJava","Version"]

# check rJava version  
.jinit()
.jcall("java/lang/System", "S", "getProperty", "java.runtime.version")
# get latest Java/Oracle version: https://www.oracle.com/technetwork/java/javase/downloads/index-jsp-138363.html  

# install relevant packages   
packages <- c("Matrix","deSolve","mvtnorm","LaplacesDemon","coda","adaptMCMC","sp","RNetLogo","ggplot2","RCurl","RColorBrewer","Interpol.T","lubridate","ggExtra","tidyr","ggthemes","reshape2","pse","sensitivity","beepr")  
if(require(packages)){
  install.packages(packages,dependencies = T)
}
# load annoying packages manually because they're stubborn 
if(mac==0){
  install.packages("RNetLogo")
  install.packages("RCurl")
  install.packages("Interpol.T")
  install.packages("lubridate")
  install.packages("tidyr")
  install.packages("ggthemes")
  install.packages("ggExtra")
  install.packages("beepr")
}
ppp <- lapply(packages,require,character.only=T)
if(any(ppp==F)){cbind(packages,ppp);cat("\n\n\n ---> Check packages are loaded properly <--- \n\n\n")}

cs <- list() # diagnostics list for checking NAs in create snails command  

################################  compile packages and load files ###################################

### Install rtools and gcc for using C code and coda package 
#### https://cran.r-project.org/bin/macosx/tools/

# define paths for gcc compiler 
if(mac==1){ #### Mac OSX
  rtools <- "/usr/local/clang6/bin"
  gcc <- paste0("usr/local/clang6/gcc-",ver_gcc,"/bin")
  # Mac OSX
  }else{ #### Windows 
  rtools <- "C:\\Rtools\\bin"
  gcc <- paste0("C:\\Rtools\\gcc-",ver_gcc,"\\bin")
}

#### point to path on comp to access rtools and gcc for C compiler  
path <- strsplit(Sys.getenv("PATH"), ";")[[1]]
new_path <- c(rtools, gcc, path)
new_path <- new_path[!duplicated(tolower(new_path))]
Sys.setenv(PATH = paste(new_path, collapse = ";"))

if(mac==1){
    # dyn.unload("IndividualModel_IBM3.so") # unLoad .so (Mac OSX
    system(paste0("R CMD SHLIB ",deb_compile,".c")) # generates .o and .so files 
    dyn.load(paste0(deb_compile,".so")) # Load .so (Mac OSX)
}else{
  # compile model from C definition
  #dyn.unload(paste0(deb_compile,".dll")) # unload dll (Windows only)
  system(paste0("R CMD SHLIB ",deb_compile,".c"))
  dyn.load(paste0(deb_compile,".dll"))# Load dll (Windows only)
}

####################################  load deb params #######################################

# load DEB starvation model parameters and create mcmc (and convert mcmc chain to coda format)
samps = readRDS(deb_samps)
samps <- as.mcmc(samps[, c("iM", "k", "M", "EM", "Fh", "muD", "DR", "fe", "yRP",
                           "ph", "yPE", "iPM", "eh", "mP", "alpha", "yEF", "LM",
                           "kd", "z", "kk", "hb", "theta", "mR", "yVE", "yEF2",
                           "sd.LI1", "sd.LU1", "sd.EI1", "sd.EU1", "sd.W1",  "sd.LI2",
                           "sd.LU2", "sd.EI2", "sd.EU2", "sd.W2", "gammaH", "gammaP", "lpost")])


# ---------- summarise and plot estimated params
svar <- "M" # select variable 
sampsvar <- samps[,svar] # pull from mcmc
summary(sampsvar) # get mean, sd, se, and quantiles for each input variable  

den <- density(sampsvar) # get AUC
densplot(sampsvar, show.obs = F,type="n") # density estimate of each variable
polygon(den, col=adjustcolor(colv,alpha=0.5),border=colv) # fill AUC 

plot(sampsvar,trace=T,density=T,col=colv) # traceplot (below) and density plot (above)
# intensive  
traceplot(sampsvar,smooth=T,type="l",lwd=0.3,xlim=c(0,length(sampsvar)),col=colv[2],xlab=paste0("Iterations"),ylab=paste0("Sampled values"),main=paste0("Sampled values over iterations for ",svar)) # iterations vs sampled valued per variable 

if(mcmcplot==1){
  par(mfrow=c(1,1))
	plotlist <- list()
	pdf("mcmc_vars.pdf",onefile = T,paper="a4")
	for(i in colnames(samps)){
	  par(bty="n", las = 1)
		if(traceplot==1){
		traceplot(sampsvar,smooth=T,type="l",xlim=c(0,length(sampsvar)),col=colv[2],xlab=paste0("Iterations"),ylab=paste0("Sampled values"),main=paste0("Sampled values over iterations for ",svar)) # iterations vs sampled valued per variable
		  }
	  svar <- i # select variable 
  	sampsvar <- samps[,svar] # pull from mcmc
  	den <- density(sampsvar) # get AUC
  	densplot(sampsvar, show.obs = F,type="n",main=paste0("Density estimate of ",i)) # density estimate of each variable
  	polygon(den, col=adjustcolor(colv,alpha=0.5),border=colv) # fill AUC 
  	}
	dev.off()
	} # end mcmcplot
# ----------

# get the best fit DEB parameters to match the data (using mcmc)
read.csv("pars.txt",header=T,sep="/",fill=T,flush=T,strip.white=T,row.names=NULL)
pars = as.vector(data.frame(samps)[max(which(data.frame(samps)$lpost >= max(data.frame(samps)$lpost) -0.001)),])
pars["Fh"] = 2 # f_scaled (for v.1.1)
pars["ENV"] = 500 # Units: L
pars["r"] = 1   # Units: day-1
pars["step"] = 1  # Units: day
pars["epsilon"] = 20 # Units: L host-1, day-1 (Rounded estimate from Civitello and Rohr)
pars["sigma"] = 0.5 
pars["m_M"] = 1   # Units: day-1
pars["m_Z"] = 1   # Units: day-1
pars["M_in"] = 10
pars["K"] = 5
pars["Det"] = 0.1 # Units mg C/L-1 d-1 (detritus)

####################################  solve deb eqs #######################################
# display list of param definitions
read.csv("debfunction.txt",header=T,sep="/",fill=T,flush=T,strip.white=T,row.names=NULL)
DEB = function(step, Food, L, e, D, RH, P, RP, DAM, HAZ, iM, k, M, EM, 
               Fh, muD, DR, yRP, ph, yPE, iPM, eh, mP, alpha, yEF,
               LM, kd, z, kk, hb, theta, mR, yVE, ENV, Lp, SAtotal, r, K, Det){
  # starting conditions 
  initials = c(Food=Food, L=L, e=e, D=D, RH=RH, P=P, RP=RP, DAM=DAM, HAZ=HAZ)
  # deb parameters
  parameters = c(iM, k, M, EM, Fh, muD, DR, yRP, ph, yPE, iPM,
                 eh, mP, alpha, yEF, LM, kd, z, kk, hb, theta, mR, yVE, ENV, Lp, SAtotal, r, K, Det)
  # estimate starting deb conditions using fitted params by solving ode's
  ## return survival and host shell length  
  DEBstep <- lsoda(initials, c(0,step), func = "derivs", dllname = deb_compile, 
                   initfunc = "initmod",  nout=2, outnames=c("Survival", "LG"), maxsteps=500000,
                   as.numeric(parameters),  rtol=1e-6, atol=1e-6, hmax=1)
  DEBstep[2, 2:12] # 12 = survival
  } # end deb model

### deb output for each timestep 
result = DEB(step=1, Food=5, L=10, e=0.9, D=as.numeric(pars["DR"]), RH=0, P=0, RP=0, DAM=0, HAZ=0, iM=pars["iM"], k=pars["k"], M=pars["M"], EM=pars["EM"], 
    Fh=pars["Fh"], muD=pars["muD"], DR=pars["DR"], yRP=pars["yRP"], ph=pars["ph"], yPE=pars["yPE"], iPM=pars["iPM"], eh=pars["eh"],
    mP=pars["mP"], alpha=pars["alpha"], yEF=pars["yEF"], LM=pars["LM"], kd=pars["kd"], z=pars["z"], kk=pars["kk"], hb=pars["hb"],
    theta=pars["theta"], mR=pars["mR"], yVE=pars["yVE"], ENV=pars["ENV"], Lp=10,SAtotal=7007.822, r=pars["r"], K=pars["K"], Det=pars["Det"]) 

### Exposure submodel
# pass the deb state vars into infection model 
Infection = function(snail.stats, miracidia, parameters){
  # Parameters
  epsilon = as.numeric(parameters["epsilon"])
  sigma = as.numeric(parameters["sigma"])
  ENV = as.numeric(parameters["ENV"])
  m_M = as.numeric(parameters["m_M"])
  step = as.numeric(parameters["step"])
  
  # Later calculations depend on exposure probabilities
  exp.rates = epsilon/ENV*(snail.stats[,"L"]>0) # This is just to get uniform exposure rates
  sum.exp.rates = sum(exp.rates)
  
  # Probabilities for fate of miracidia
  ## Still in water
  P.left.in.water = exp(-(m_M+sum(exp.rates))*step)
  ## Infect a snail
  P.infects.this.snail = (1 - P.left.in.water)*(sigma*exp.rates/(m_M+sum.exp.rates)) 
  ## Die in water or fail to infect
  P.dead = (1 - P.left.in.water)*(m_M/(m_M+sum.exp.rates)) + sum((1 - P.left.in.water)*((1-sigma)*exp.rates/(m_M+sum.exp.rates)))
  
  prob.vector = c(P.infects.this.snail, P.left.in.water, P.dead)
   
  # Multinomial outcome from number of miracidia in env based on their survival probability
  rmultinom(n=1, size=miracidia, prob=prob.vector)
  #sum(P.left.in.water, P.invades.this.snail, P.dead)
} # end infection model 

### update all the snails @netlogo
update.snails = function(who, new.L, new.e, new.D, new.RH, new.P, new.RP, new.DAM, new.HAZ, new.LG){
  paste("ask snail", who, 
        "[set L", new.L,
        "set ee", new.e,
        "set D", new.D,
        "set RH", new.RH,
        "set P", new.P,        
        "set RPP", new.RP,       
        "set DAM", new.DAM,
        "set HAZ", new.HAZ,
		"set LG", new.LG, # new max length
        "]")
} # end host update

#Example update
#paste(mapply(update.snails, who=snail.stats[,"who"], new.L=L, new.e=e, new.D=D, new.RH=RH, new.P=P, new.RP=RP, new.DAM=DAM, new.HAZ=HAZ), collapse=" ")

geterrmessage() # check if there were any error messages  

###########################################################################################
####################################  load netlogo ######################################## 
###########################################################################################
# @netlogo

# working NLStart in RStudio. works with gui=F (2018/09/24)
if(gui==0){
  NLStart(nl.path,gui=F,nl.jarname = paste0("netlogo-",ver_nl,".jar")) # open netlogo without a gui  
  }else{
    NLStart(nl.path,nl.jarname = paste0("netlogo-",ver_nl,".jar")) # open netlogo
  }

NLLoadModel(paste0(model.path,nl.model),nl.obj=NULL) # load model  
# if java.lang error persists on Mac, try copying all .jar files from the 'Java' folder where Netlogo is installed into the main Netlogo folder   	

################################################################################################
####################################  start netlogo sim ######################################## 
################################################################################################

# set type of resource input @netlogo
set_resource_type<-function(resource_type){ # set resource input in env  
  if(resource_type == "detritus"){NLCommand("set resource_type \"detritus\" ")}else{NLCommand("set resource_type \"algae\" ")}}
set_resource_type(resource_type) # set resource type: "detritus" or "algae"  @netlogo

# set type of resource dynamics @netlogo
set_resources<-function(resources){ # set resource input in env  
  if (resources == "cyclical"){NLCommand("set resources \"cyclical\" ")}else{NLCommand("set resources \"event\" ")}}
set_resources(resources) # set resources: "cyclical" or "event"  @netlogo
cat("\nResource type = ",resource_type,"\nResources = ",resources)

testrun <- 0 # do a quick testrun to see plots
snail_control <- 0 # 1 = add molluscicide event

if(save_to_file==1){pdf(paste0(wd,"/master_sim.pdf"),onefile=T,paper="a4")}
ifelse(testrun==1,n.ticks<-20,n.ticks<-120)

# snail control params
me_pars <- seq(10,110,10) # molluscicide events (me)
me_90 <- 2.3 # background hazard rate for 90% snail mortality from molluscicide event (per day) 

# algae params
rg_pars <- c(0.5,0.75,1) # resource growth rates (rs)
alpha_pars <- c(0,0.25,0.5,0.75,1) # amplitude of resources (alphas)
rho_pars <- c(1,seq(10,120,10)) # periodicity of resources (rhos)

# detritus params
detr_pars <- c(0.25,0.35,0.5) # detritus input (mg L^-1 day^-1)

# mortality params 
hb_pars <- c(0.0001, 0.001, 0.002, 0.004, 0.006, 0.008, 0.01, 0.02, 0.03, 0.04); hb_pars

me_pars <- 0 # set molluscicide events to 0 
if(resource_type=="detritus"){
  detr_pars <- detr_pars;alpha_pars <- 0; rho_pars <- 10; rg_pars <- 0;cat("detritus input = ",detr_pars,"\nrg = ",rg_pars,"\nhb = ", hb_pars)
  }else{detr_pars <- 0;cat("detritus input = ", detr_pars,"\nrg = ",rg_pars)}
# set resource to cycle or be constant
if(resource_type=="algae"){
  if(resources=="cyclical"){alpha_pars <- alpha_pars; rho_pars <- rho_pars ; rg_pars <- rg_pars;cat("alphas = ",alpha_pars,"\nrhos = ",rho_pars,"\nrgs = ",rg_pars)
  }else{alpha_pars <- 0; rho_pars <- 10; rg_pars <- rg_pars;cat("alphas = ",alpha_pars,"\nrhos = ",rho_pars,"\nrgs = ",rg_pars,"\nhb = ", hb_pars)}}

Env_G = numeric() # create empty environment vector 
if(snail_control==1){me_pars <- me_pars}else{me_pars <- 1000000}; cat("Snail control will occur every ",max(me_pars)/me_pars[1]-1," days") 

# individual outputs
cerc_list <- list() # cercariae   
food_list <- list() # food in env 
juv_list <- list() # juvenile hosts
adult_list <- list() # adult hosts 
infec_list <- list() # infected hosts
infec_shed_list <- list() # infected shedding hosts
hl_list <- list() # host length
pmass_list <- list() # parasite biomass 

# master outputs
cerc_master <- list() # master list for cerc density (Env_Z) 
food_master <- list() # master list for food dynamics (Env_F) 
juv_master <- list() # master list for total host pop () 
adult_master <- list() # master list for total host pop () 
infec_master <- list() # master list for infected host pop () 
infec_shed_master <- list() # master list for infected shedding host pop
hl_master <- list() # master list for host length
pmass_master <- list() # master list for parasite biomass 

# define plot window
plot.matrix <- matrix(c(length(alpha_pars),length(rho_pars)))
par(mfrow=plot.matrix)

####################################  start netlogo sim ######################################## 
for(hb in hb_pars){
  for(detr in detr_pars){ # loop through detritus inputs 
    for(alpha in alpha_pars){ # loop through alphas (amplitude in food cycle)
      for(rho in rho_pars){ # loop through rhos (periodicity of food cycle)
    	  for(rg in rg_pars){ # loop through rgs (food growth rates)
    	    for(me in me_pars){ # loop through mes (molluscicide events)
    	      NLCommand("setup")
    	      for(t in 1:n.ticks){ # start nl sim  @netlogo
    	        snail.stats = NLGetAgentSet(c("who", "L", "ee", "D", "RH", "P", "RPP", "DAM", "HAZ", "LG"), "snails")
              N.snails = length(snail.stats[,"L"])
              environment = as.numeric(NLGetAgentSet(c("F", "M", "Z", "G"), "patches")) # calc food, free miracidia, cercariae released, and eggs, per patch
        
              # Infect snails
              Infection.step = as.vector(Infection(snail.stats, environment[2], pars)) # Who gets infected
              snail.stats[which(Infection.step[1:N.snails] > 0),"P"] = snail.stats[which(Infection.step[1:N.snails] > 0),"P"] + 2.85e-5 # add biomass of one miracidia
              
              # define food dynamics for cyclical algal (logistic food growth equation) or detritus food sources
              alpha <- alpha # amplitude of resources
              rho <- rho  # periodicity (time range of resource cycles)  
              rg <- rg # resource growth rate 
              rg_t <- rg + alpha * rg * sin(2 * pi * t/rho) # equilibrium cyclical resource dynamics (19-12-18)
              pars["r"] <- rg_t # set resource growth rate 
              pars["Det"] <- detr # Units mg C/L-1 d-1 (detritus)
              # Update DEBS, HAZ=0 so survival probs are calculated for the current day
              snail.update = t(mapply(DEB, L=snail.stats[,2], e=snail.stats[,3], D=snail.stats[,4], RH=snail.stats[,5],
                                      P=snail.stats[,6], RP=snail.stats[,7], DAM=snail.stats[,8], Lp=snail.stats[,10],# Food=environment[1]*(snail.stats[,2]^2)/sum(snail.stats[,2]^2), # update food availability per snail 
                                      MoreArgs = list(step=1, HAZ=0, Food=environment[1],# constant food available (23-1-19)
                                                      iM=pars["iM"], k=pars["k"], M=pars["M"], EM=pars["EM"], Fh=pars["Fh"], 
                                                      muD=pars["muD"],
                                                      DR=pars["DR"], yRP=pars["yRP"], ph=pars["ph"], yPE=pars["yPE"], iPM=pars["iPM"], eh=pars["eh"],
                                                      mP=pars["mP"], alpha=pars["alpha"], yEF=pars["yEF"], LM=pars["LM"], kd=pars["kd"], z=pars["z"], 
                                                      kk=pars["kk"], 
                                                      #if(snail_control==1){
                                                      # if(day==me){hb <- me_90}
                                                      # }else{hb <- pars["hb"]},
                                                      hb=hb, # hb pars test
                                                      theta=pars["theta"], mR=pars["mR"], yVE=pars["yVE"], SAtotal= sum(snail.stats[,2]^2), 
                                                      ENV=pars["ENV"], r=pars["r"], K=pars["K"], 
                                                      Det=pars["Det"]))) # detritus (Det) defined in C file
              L = snail.update[,"L"] # host structural length
              e = snail.update[,"e"] # host scaled reserve density    
              D = snail.update[,"D"] # host development 
              RH = snail.update[,"RH"] # host energy to reproduction buffer  
              DAM = snail.update[,"DAM"] # host damage from starvation  
              HAZ = snail.update[,"HAZ"] # host hazard rate from starvation   
              LG = snail.update[,"LG"] # host shell length  
              P = snail.update[,"P"] # parasite mass (sum within host)
              RP = snail.update[,"RP"] # parasite reproductive buffer  
              # ingestion = environment[1] - sum(snail.update[,"Food"]) # food intake by host from environment (for v.1.1)
              hl_list[t] <- L # get host lengths per model step
              pmass_list[t] <- P # get parasite mass per model step 
              
              Eggs = floor(RH/0.015)  # Figure out how many (whole) eggs are released  
              # if(day==me){Eggs <- Eggs[1:round(0.1*length(Eggs))]} # kill off 90% of snail eggs in water with molluscicide event  
              RH = RH %% 0.015        # Remove released cercariae from the buffer
              Cercs = floor(RP/4e-5)  # Figure out how many (whole) cercs are released
              RP = RP %% 4e-5         # Remove released cercariae from buffer
              Eggs = as.integer(Eggs); Cercs = as.integer(Cercs)
              
              # Update environment 
              Env_M = as.numeric(Infection.step[N.snails + 1] + pars["M_in"]) # total miracidia density 
              Env_Z = as.numeric(environment[3]*exp(-pars["m_Z"]*pars["step"]) + sum(Cercs)/pars["ENV"]) # total cerc density
              Env_G = as.integer(Env_G) # set pop density outputs to integer to pass into Env_G and rbinom func
              # ifelse(day==me,Env_G[day] <- max(0, 0.1*sum(Eggs),na.rm=T),Env_G[day] <- max(0, sum(Eggs),na.rm=T)) # kill off 90% of snail eggs in water with molluscicide event 
              Env_G[day] <- max(0, sum(Eggs),na.rm=T)
              
              Env_G[is.na(Env_G)] <- 0 # turn NAs to 0 to feed into rbinom function  
              # Env_F = max(0.001, as.numeric(pars["K"]*environment[1]/(environment[1] + (pars["K"] - environment[1])*exp(-pars["r"]*pars["step"])) - ingestion)) # Analytical soln to logistic - ingestion (alphas [1,100]) (original r growth equation)
              # Env_F = max(0.001, as.numeric(pars["K"]*environment[1]/(environment[1] + (pars["K"] - environment[1])*exp(-rg_t*pars["step"])) - ingestion)) # Analytical soln to logistic - ingestion with equilibrium resource growth wave (rg_t) (alphas [0,1]) (for v.1.1)     
              # F = F * exp(- r + alpha * r * sin(2 * pi * t/rho) * s) * (1 - F/K) - f(i_{M} * sum(L^2) # v. 1.2 algae and detritus with cyclical algal growth
              # r_t <- pars["r"] + alpha * pars["r"] * sin(2 * pi * t/rho) # equilibrium resource dynamics (static)
              Env_F = max(0.001, snail.update[1,"Food"]) # algal or detritus food sources (for v.1.2)
              # Command back to NL @netlogo
              NLCommand("ask patch 0 0 [set F", Env_F, "set M", Env_M, "set Z", Env_Z, "set G", Env_G[day], "]")
              snail.commands = paste(mapply(update.snails, who=snail.stats[,"who"], new.L=L, new.e=e, new.D=D, new.RH=RH, new.P=P, new.RP=RP, new.DAM=DAM, new.HAZ=HAZ, new.LG=LG), collapse=" ")
              NLCommand(snail.commands) 
              if(day > 10){
                if(snail_control==1){ # kill off 90% of snail eggs in water with molluscicide event  
                  if(day==me){create_snails <- rbinom(n=1, size=Env_G[day - 10], prob=0.1)}
                }else{create_snails <- rbinom(n=1, size=Env_G[day - 10], prob=0.5)}
                NLCommand("create-snails ", create_snails, "[set L 0.75 set ee 0.9 set D 0 set RH 0 set P 0 set RPP 0 set DAM 0 set HAZ 0 set LG 0.75]")
              } # end create snails
              NLCommand("go") # run @netlogo sim steps
              #cs[t] <- rbinom(n=1, size=Env_G[day - 10], prob=0.5) # list to check 'create snails' output doesn't produce NAs
              day = day + 1 
              if(testrun==1){
                cerc_list[t] <- Env_F + rho # use to test plot outputs quickly (plots food + rho value as mock output to show amplitude)  
                }else{
                  # results outputs
                  cerc_list[t] <- Env_Z # get cercariae density 
      		        food_list[t] <- Env_F # get food growth
      		        juv_list[t] <- length(which(snail.stats$RH==0)) # get juvenile hosts
      		        adult_list[t] <- length(which(snail.stats$RH>0)) # get adult hosts
      		        infec_list[t] <- length(which(snail.stats$P>0)) # get just infected hosts
      		        infec_shed_list[t] <- length(which(snail.stats$RP>0)) # get infected hosts that are shedding
      		        } # end testrun
              } # --------------------------------------- end nl sim
    	      # save individual outputs 
    	      cerc_list <- as.numeric(cerc_list) 
      	    food_list <- as.numeric(food_list)
      	    juv_list <- as.numeric(juv_list)
      	    adult_list <- as.numeric(adult_list)
            infec_list <- as.numeric(infec_list)
            infec_shed_list <- as.numeric(infec_shed_list)
            hl_list <- as.numeric(hl_list)
            pmass_list <- as.numeric(pmass_list)
            # save master outputs 
            cerc_master[[length(cerc_master)+1]] <- cerc_list # cerc master list
            food_master[[length(food_master)+1]] <- food_list # food master list
            juv_master[[length(juv_master)+1]] <- juv_list # juv pop master list
            adult_master[[length(adult_master)+1]] <- adult_list # adult pop master list
            infec_master[[length(infec_master)+1]] <- infec_list # infected host pop master list
            infec_shed_master[[length(infec_shed_master)+1]] <- infec_shed_list # infected shedding host pop master list
            hl_master[[length(hl_master)+1]] <- hl_list # host length master
            pmass_master[[length(pmass_master)+1]] <- pmass_list # host length master
            if(save_to_file==1){dev.off()}
            } # --------------- end mes
    	    } # ------------------------------ end rgs
    	  } # --------------------------------------------- end rhos
      } # ----------------------------------------------------------- end alphas
    } # ------------------------------------------------------------------------- end detritus
} # end hb pars 
####################################  end netlogo sim ######################################## 

# results output 
# save sim results to dir 
str(list(cerc_master,food_master,juv_master, adult_master,infec_master,infec_shed_master,hl_master,pmass_master))  
global_output <- list(cerc_master,food_master,juv_master, adult_master,infec_master,infec_shed_master,hl_master,pmass_master) 
saveRDS(global_output,paste0(wd,"/global_output_",resource_type,"_",resources,"_hbs.R"))
# read in saved sim results
cat("order = cerc, food, juv, adult, infected, infected shedding, host length, parasite mass")
cat("detritus =  ",c(0.25,0.35,0.5))
cat("algae with rg = ",c(0.5,0.75,1))

# NLQuit()

#################################################################################################
##########################################  end body ############################################ 
#################################################################################################
```

End Appendix  